Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-07 Manuscript/manuscript.Rmd
editted abstract

**Background.** Chromosome conformation capture sequencing technologies have shown that the 3-dimensional structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs). The locations of TAD boundaries have been shown to be highly conserved. Additionally, the distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches toward identifying the epigenomic features most predictive of TAD boundaries.

Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions on the linear genome that include TAD boundaries. However, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced classes distinguished by boundary vs. non-boundary regions and the use of distances ($log_{2}$ transformed) as the best representation of the association between TAD boundaries and genomic annotations.

**Methods.** Using replicate Hi-C data for the GM12878 and K562 cell lines, we proposed a novel ensemble framework that addresses these data irregularities and characteristics. Multiple random forest classification models built on regularized sets of genomic annotations obtained from the UCSC genome browser database, in combination with variations in data resolution, feature engineering, and re-sampling techniques. The composite metric Matthew's Correlation Coefficient (MCC) was used to compare models derived from the framework to determine a consensus approach toward predicting TAD boundary formation.

**Results.** We discovered that models built on SMOTE (Synthetic Minority Over-sampling TEchnique) re-sampled data out performed models built simply on random under- and over-sampled data. Additionally, our results indicated that log2 distances between annotations and TAD boundaries were more informative than other predictor-types used to build models. Importantly, we also found that performances tended to degrade as data resolution decreased. Lastly, the most predictive annotations in the best performing models included known architectural proteins such as CTCF, SMC3, RAD21, and ZNF143, as well as lesser known transcription factors including P300 and NFKB.

**Conclusions.** The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction.

>>>>

**Background.** Advances in chromosome conformation capture sequencing technologies have revealed the hierarchical structure of genome organization. Among this organization exists regions in the genome that are highly self-interacting, termed topologically associated domains (TADs). The locations of TADs throughout the linear genome has been studied extensively. However, the set of functional genomic annotations that are associated with the formation of TADs has not been systematically evaluated. By identifying the annotations that are enriched near regions that demarcate TADs, we can gain a better understanding of the mechanisms that lead to the 3D structure of the genome.

This is a multi-faceted problem, and many current methods ignore key characteristics of domain data when building models and reporting performances. First, the sparsity of TAD boundaries throughout the linear genome leads to heavily imbalanced classes distinguished by boundary vs. non-boundary regions. Second, determining the most informative relationship between TAD boundaries and genomic annotations remains and open problem. We hypothesize that implementing re-sampling, in conjunction with distances (in basepairs) between annotations and pre-defined regions on the genome, will results in better learned models.

**Methods.** We proposed a novel ensemble framework in order to systematically compare the performances of random forest classifiers implemented on combinations of different Hi-C contact matrix resolutions, feature engineering strategies, and re-sampling techniques.

**Results.** Using the GM12878 and K562 cell lines, we show that models built on SMOTE (Synthetic Minority Over-sampling TEchnique) re-sampled data out performed models built simply on random under- and over-sampled data, across all resolutions. Additionally, our results indicated that distances between annotations and TAD boundaries were more informative than other predictor-types used to build models. Importantly, we also found that performances tended to degrade as data resolution decreased. Lastly, the most predictive annotations in the best performing models were found to be consistent across both cell line and resolution. They included well-known marks of TAD boundaries such as CTCF, SMC3, RAD21, and ZNF143, as well as lesser known transcription factors including P300 and NFKB.

**Conclusions.** The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Our framework provides insight into the best set of parameter necessary for optimal performance.
